#Reproducible Research Peer Assessment 1
  
## Loading and preprocessing the data
Read **activity.csv** file into variable *activity*.  
Convert the date column into a YYYY-MM-DD format.
````{r, echo=TRUE}
activity <- read.csv("activity.csv", sep=",", header=TRUE, stringsAsFactors=FALSE, na.strings=NA)
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
head(activity); tail(activity)
```
Load libraries *dplyr*, *ggplot2*
````{r, echo=TRUE, results='hide'}
library(dplyr)
library(ggplot2)
```
## Mean total number of steps taken per day
Obtain the total number of steps each day by grouping by day, then finding the total number of steps for that day.  
Plot of histogram showing the distribution of the total steps each day.  
The data excludes the NA values.
```{r, echo=TRUE}
steps.day <- activity %>%
  group_by(date) %>%
  summarise(sum(steps, na.rm=TRUE))
colnames(steps.day) <- c("date", "total.steps")
total.steps.plot <- ggplot(steps.day, aes(x=total.steps)) +
  geom_histogram(binwidth=800) +
  ggtitle("Histogram of Total Steps per day")
total.steps.plot
```
  
Mean and median steps taken excluding NA values
```{r, echo=TRUE}
steps.summary <- as.data.frame(cbind(mean(steps.day$total.steps), median(steps.day$total.steps)))
colnames(steps.summary) <- c("mean", "median")
head(steps.summary); tail(steps.summary)
````
## Average daily activity pattern
```{r, echo=TRUE}
mean.steps <- activity %>%
  group_by(date) %>%
  summarise(mean(steps, na.rm=TRUE))
mean.steps <- subset(mean.steps[which(complete.cases(mean.steps)==TRUE),])
colnames(mean.steps) <- c("date", "mean.steps")
meansteps.plot <- ggplot(mean.steps, aes(x=date, y=mean.steps)) +
  geom_line() +
  ggtitle("Mean Number of Steps Per Day") + xlab("Date") + ylab("Mean number of steps")
meansteps.plot
```
### Mean number of steps by 5-min interval
Obtain *steps.int* which summarizes the mean steps taken at each time interval across all subjects.
```{r, echo=TRUE}
steps.int <- activity %>%
  group_by(interval) %>%
  summarise(mean(steps, na.rm=TRUE))
colnames(steps.int) <- c("interval", "mean.steps")
```
5-min interval with the highest number of steps
```{r, echo=TRUE}
steps.int[which(steps.int$mean.steps==max(steps.int$mean.steps)),]
```
## Imputing missing values
Remove NA values from the steps column by using an intermediate variable *act.na* by:  
1. subsetting *act.na* from the *activity* dataset containing only NA values in the steps column.  
2. using the *left_join* function, merge the *act.na* and *steps.int* datasets  
3. impute the missing (NA) values by selecting the mean steps at each 5-min interval from the mean.steps column.  
4. *anti_join* dataset *act.na* to the main *activity* dataset to eliminate the copies of rows with missing values (NA) in the steps column into a new variable *act.imp*.  
5. rbind *act.na* and *act.imp* into *act.imp* to obtain a complete dataset with imputed values
```{r, echo=TRUE}
act.na <- filter(activity, is.na(steps)==T) #1
act.na <- left_join(act.na, steps.int, by="interval") #2
act.na <- transmute(act.na, steps = mean.steps, date=date, interval=interval) #3
act.imp <- anti_join(activity, act.na, by=c("date", "interval")) #4
act.imp <- rbind(act.na, act.imp) %>% #5
  arrange(date, interval)
```
### Number of missing values in dataset
```{r, echo=TRUE}
length(act.na$steps)
```
### Histogram of adjusted total steps each day
```{r, echo=TRUE}
steps.day.imp <- act.imp %>%
  group_by(date) %>%
  summarise(sum(steps, na.rm=TRUE))
colnames(steps.day.imp) <- c("date", "total.steps")
total.steps.plot.imp <- ggplot(steps.day.imp, aes(x=total.steps)) +
  geom_histogram(binwidth=800) +
  ggtitle("Histogram of Total Steps per day\n(Adjusted)")
total.steps.plot.imp
```
  
### Adjusted mean and median total steps each day
```{r, echo=TRUE}
steps.summary.imp <- as.data.frame(cbind(mean(steps.day.imp$total.steps), median(steps.day.imp$total.steps)))
colnames(steps.summary.imp) <- c("mean", "median")
steps.summary.imp
```
## Differences in activity patterns between weekdays and weekends
Create a new *day.type* column in the *act.imp* (activity with imputed values) dataset indicating either weekend or weekday
```{r, echo=TRUE}
act.imp <- mutate(act.imp, date, day.type=ifelse(weekdays(act.imp$date)%in%c("Saturday", "Sunday"), "Weekend", "Weekday"))
act.imp$day.type <- as.factor(act.imp$day.type)
```
Mean number of steps each day
```{r, echo=TRUE}
daytype.comp <- act.imp %>% 
  group_by(interval, day.type) %>%
  summarise(mean.steps=mean(steps))
head(daytype.comp); tail(daytype.comp)
```
Time series plot of mean steps taken each day by weekdays and weekends
```{r, echo=TRUE}
daytype.plot <- ggplot(daytype.comp, aes(x=interval, y=mean.steps)) +
  geom_line() + facet_grid(. ~ day.type) +
  ggtitle("Time Series Plot of Mean Steps Taken") +
  xlab("Interval (mins)") + ylab("Mean Number of Steps")
daytype.plot
```