---
title: "Reproducible Research Peer Assessment 2"
author: "Jansen Ng"
output: html_document
---
### Read files and load libraries  
- The *repdata-data-StormData.csv* file is read into the variable *stormData*  
```{r, cache=TRUE, echo=TRUE, results='hide'}
stormData <- read.table(file="repdata-data-StormData.csv", sep=",", stringsAsFactors=FALSE, header=TRUE, na.strings=c("", "NA"))
```

- The packages *dplyr*, *ggplot2*, *reshape2* are loaded
```{r}
library(dplyr)
library(reshape2)
library(ggplot2)
```

## Most Dangerous to Population Health

### Data Processing  
- The 2 variables which contain information about harm or damage to human health are *FATALITIES* and *INJURIES*.  
- A subset of the *stormData* dataset containing *EVTYPE*, *FATALITIES*, *INJURIES* is created, then the sum of fatalities and injuries are aggregated by evtype.  
- The resulting data frame is then separately arranged in descending order of total fatalities and total injuries.  
```{r}
popHealth <- stormData %>% select(EVTYPE, FATALITIES, INJURIES)
popHealth$EVTYPE <- as.factor(popHealth$EVTYPE)
ph1 <-  popHealth %>% group_by(EVTYPE) %>%
          summarize(total.fatalities=sum(FATALITIES), total.injuries=sum(INJURIES))
ph1.fatalities <- ph1 %>% arrange(desc(total.fatalities))
ph1.injuries <- ph1 %>% arrange(desc(total.injuries))
```

### Results  
#### Highest total fatalities  
The top 5 storm event types are as follows:  
```{r}
head(ph1.fatalities, n=5)
```

#### Highest total injuries  
The top 5 storm event types are as follows:  
```{r}
head(ph1.injuries, n=5)
```

### Conclusion  
As the storm event type tornado gives the highest total fatalities as well as total injuries, the storm event type most dangerous to human health is the tornado.

## Greatest Economic Consequences across the United States 

### Data Processing  
- The variables that would be relevant to this question are *STATE*, *EVTYPE*, *PROPDMG*, *PROPDMGEXP*, *CROPDMG*, *CROPDMGEXP*.
- Flood-related damage under the variable *WFO* are excluded, and NAs are regarded as 0 in terms of damage costs.
```{r}
ecDmg <- stormData %>% select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
ecDmg$EVTYPE <- as.factor(ecDmg$EVTYPE)
ecDmg <- ecDmg %>% mutate(
  p.multip=ifelse(PROPDMGEXP=="B", 1e+09, ifelse(PROPDMGEXP=="M", 1e+06, ifelse(PROPDMGEXP=="K", 1000, NA))), 
  c.multip=ifelse(CROPDMGEXP=="B", 1e+09, ifelse(CROPDMGEXP=="M", 1e+06, ifelse(CROPDMGEXP=="K", 1000, NA))))
```

#### Damage to Crops  
Create dataset of storm event type in decreasing magnitude of damage to crops  
```{r}
ecDmg.c <- ecDmg %>% select(EVTYPE, CROPDMG, c.multip) %>%
  filter(is.na(c.multip)==FALSE) %>%
  mutate(crop.dmg=CROPDMG*c.multip) %>%
  select(EVTYPE, crop.dmg) %>%
  group_by(EVTYPE) %>%
  summarize(total.crop.dmg=sum(crop.dmg)) %>%
  arrange(desc(total.crop.dmg))
```

#### Damage to Property  
Create dataset of storm event type in decreasing magnitude of damage to property  
```{r}
ecDmg.p <- ecDmg %>% select(EVTYPE, PROPDMG, p.multip) %>%
  filter(is.na(p.multip)==FALSE) %>%
  mutate(prop.dmg=PROPDMG*p.multip) %>%
  select(EVTYPE, prop.dmg) %>%
  group_by(EVTYPE) %>%
  summarize(total.prop.dmg=sum(prop.dmg)) %>%
  arrange(desc(total.prop.dmg))
```

### Results  
Subset top 10 events to property and crops
```{r}
cDmgTop10 <- subset(ecDmg.c[1:10,])
pDmgTop10 <- subset(ecDmg.p[1:10,])
```

#### Most Damage to Crops

```{r, results='hold', fig.width=11}
cDmgTop10

ggplot(cDmgTop10, aes(x=factor(EVTYPE), y=total.crop.dmg)) +
  geom_bar(stat="identity") +
  ggtitle("Top 10 Most Damage to Crops") +
  ylab("Total Damage ($)") + xlab("Event Type")
```

#### Most Damage to Property

```{r, results='hold', fig.width=11}
pDmgTop10

ggplot(pDmgTop10, aes(x=factor(EVTYPE), y=total.prop.dmg)) +
  geom_bar(stat="identity") +
  ggtitle("Top 10 Most Damage to Property") +
  ylab("Total Damage ($)") + xlab("Event Type")
```

### Conclusion  
Based on the above plots and the data for the damage to property and crops, the most devastating storm event to the economy is the flood.